---
title: "Wage Labour and Capital"
output: 
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE }
library(dplyr)
library(ggplot2)
library(tidyr)
library(sandwich)
library(lmtest)
library(patchwork)
library(MASS)
library(brms)
```

In this project, we translated the theoretical framework of Marx's Wage Labour and Capital into a mathematical model.
To ensure that the variables behave according to the theoretical mechanisms describes in the literature, rather than being confounded by the complexity of real historical processes, **we rely on synthetic data rather than empirical historical data. This allow us to control the data generating process precisely.** 

This simulation is structured around two independent variables : **wage gap** and **labor surplus** whose dynamics are governed by theoretically motivated equations developed in the following sections. These variables feed into **cumulative pressure terms** ($CP_w$, $CP_l$), which in turn predict the probability of **structural crisis** through a logistic regression model.(+DAG)

We observed N societies over T decades. To determine the sample size, we used the 10 events per variable (EPV) rule of thumb for logistic regression. But two methodological challenges must be addressed. First, the determination of N and T depends on the **crisis probability and the autocorrelation structure of the cumulative pressure variables.** Second, the cumulative pressure terms exhibit an **ARMA (1,q)** autocorrelation structure, making analytical derivation of the effective sample size (ESS) inappropriate. For these reasons, we proceed in two stages: a pilot simulation with provisional parameters to empirically **estimate the autocorrelation structure**, followed by the full simulation with final **N and T derived from both the empirical autocorrelation estimate and the sensitivity analysis on crisis probability.**

Natural order : estimation of the empirical autocorrelation from the pilot, derive T with the effective sample size (ESS) and N with the events per variable (EPV).

#### The Variables : 

A structural crisis occurrence at decade t can be translated in a logistic regression with the possible output : *crisis, no crisis.*

$P(crisis[t]) = logistic(z)$, with :

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$

- $x_1$: The wage gap measures the distance between the wage and the subsistence cost.

- $x_2$: The labor surplus is the excess of workers relative to available job

- $x_1x_2$ is the interaction term. It captures the idea that the effect of one variable depends on the level of the others. So when wage gap and labor surplus are high simultaneously, the combine effect on crisis probability is greater than their individual effects added together. The interaction term formalizes Marx's argument that structural crisis emerges not from wage depression or labor surplus independently, but from their simultaneous occurrence — consistent with the threshold mechanism described in Wage Labour and Capital.

Following the **Events Per Variable rule**, which recommends a minimum of 10 observed events per predictor to ensure reliable parameter estimation in logistic regression, and given that our model contains 3 variables ($x_1$, $x_2$, $x_1x_2$), the minimum number of **expected crisis events is 3×10=30**

#### Data Generating Process : Pilot Simulation

```{r}
set.seed(123)
```

```{r}
N_pilot <- 10
T_pilot <- 200
```

#### Generate wage gap and labor surplus

Labor surplus has to be generated before wage gap because the growth rate $r_w$ of the logistic trend of the wage gap depends on labor surplus

1.    Labor surplus : 

$$
LaborSurplus(t) = \frac{K_l}{1+e^{-r_l(t-t_{0,l})}} + A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P} + \epsilon
$$

```{r}
# labor_surplus parameters
K_l <- 1
r_l <- .05
t0_l  <- 50
A <- .2
P <- 5
delta_l <- .43
```

```{r}
func_labor_surplus <- function(N,T) {
  
  labor_surplus <- matrix(NA, nrow = T , ncol = N)
  for (n in 1:N) {
    for (t in 1:T) {
    
      logistic_trend <- K_l/(1+exp(-r_l*(t-t0_l)))
      oscillation <- A * exp(-delta_l*t) * sin((2*t*pi)/P)
      epsilon <- rnorm(1, mean = 0, sd = .02)
    
      labor_surplus[t, n] <- logistic_trend + oscillation + epsilon
    }
  }
  return(labor_surplus)
}
```


```{r}
labor_surplus <- func_labor_surplus(N_pilot, T_pilot)
head(labor_surplus)
```

2.    Wage gap : 


$$
WageGap(t) = \frac{K_w}{1+e^{-r_w(t-t_{0,w})}}
$$

$$
r_w = r_{base,w} + \alpha(LaborSurplus_{t-1})
$$

```{r}
# wage_gap parameters
K_w <- 1
r_base <- .03
alpha <- .5
t0_w <- 60
```

```{r}
func_wage_gap <- function(N, T, lab){
  
  wage_gap <- matrix(NA, nrow = T , ncol = N)
  for (n in 1:N) {
    for (t in 1:T) {
    
      if (t == 1) {r_w <- r_base} else {r_w <- r_base + alpha*(lab[t-1, n])}
    
     wage_gap[t,n]<- K_w/(1+exp(-r_w*(t-t0_w)))
    }
  }
  return(wage_gap)
}
```

```{r}
wage_gap <- func_wage_gap(N_pilot, T_pilot, labor_surplus)
head(wage_gap)
```

The logistic formula of the wage gap is deterministic given the growth rate $r_w$ and the time $t$. 
At $t=1$, $r_w = r_{base}$ because there is no labor surplus before $t=1$. Therefore, the wage gap at $t=1$ is identical for all societies, meaning that all societies start with the same structural conditions but diverge as the market history evolves. 

#### Generate cumulative wage gap and cumulative labor surplus
Weights are assigned inversely to chronological distance, prioritizing recent events over earlier ones via an exponential decay factor.

$$
CP\_L(t) = \sum^{t-1}_{k} L_{t-k}\cdot e^{-\delta_c\cdot k}
$$

$$
CP\_W(t) = \sum^{t-1}_{k} W_{t-k}\cdot e^{-\delta_c\cdot k}
$$

```{r}
# decay parameter (cumulative pressure)
delta_c <- 0.43
```

```{r}
cp <- function(N, T, lab, wag) {
  
  CP_l <- matrix(NA, nrow = T , ncol = N)
  CP_w <- matrix(NA, nrow = T , ncol = N)

  for (n in 1:N){
    for (t in 1:T){
    
    # cumulative pressure is only meaningful at t=2
      if (t==1){
        CP_l[t, n] <- 0
        CP_w[t, n] <- 0
      }
      else {
      
        cpsum_l <- 0
        cpsum_w <- 0
      
        for (k in 1:(t-1)){
          cpsum_l <- cpsum_l + lab[t-k, n] * exp(-delta_c*k)
          cpsum_w <- cpsum_w + wag[t-k, n] * exp(-delta_c*k)
      }
      
        CP_l[t, n] <- cpsum_l
        CP_w[t, n] <- cpsum_w
      }
    }
  }
  return( list(CP_l = CP_l, CP_w = CP_w) )
}

```

```{r}
result <- cp(N_pilot, T_pilot, labor_surplus, wage_gap)
CP_w <- result$CP_w
CP_l <- result$CP_l
```

```{r}
head(CP_l)
```

```{r}
head(CP_w)
```

-   The identical wage gap at $t=2$ across all societies can be explain by the identical wage gap at $t=1$. At $t=2$, CP_w looks back one period, meaning it only uses data in $t=1$.

```{r}
n <- 70
s <- 50
trend_l <- K_l / (1 + exp(-r_l * (s:n - t0_l)))
 
par(mfrow = c(1,2))

plot(s:n, labor_surplus[s:n, 1], type = "n",
     xlab = "Decades", ylab = "Labor Surplus",
     main = "Labor Surplus-Society 1")
lines(s:n, labor_surplus[s:n, 1], col = "#00070d", lwd = 1.5)
lines(s:n, trend_l, col = "#E24B4A", lty = 2, lwd = 2)
legend("topleft",
       legend = c("Labor surplus", "Logistic trend"),
       col = c("#00070d", "#E24B4A"),
       lty = c(1, 2, 1, 1), lwd = 2)

plot(1:T_pilot, CP_l[,1], type = "l", 
     xlab = "Decades", ylab = "Cumulative Pressure",
     main = "Cumulative Pressure-Society 1")
lines(1:T_pilot, CP_w[,1],col ="red")
legend("bottomright", legend = c("CP_l", "CP_w"), col = c("black", "red"), lty = 1)

par(mfrow = c(1, 1))
```

"Wages will now rise, now fall,according to the relation of supply and demand, according as competition shapes itself between the buyers of labour-power, the capitalists, and the sellers of labour-power, the workers. "

The first plot illustrates the labor surplus of Society 1 from decade 50 (pre-inflection) to decade 70 (post-inflection), with the dotted line representing the logistic trend. The second plot displays the cumulative wage gap and labor surplus across the $T_{pilot}$

**Phases of labor surplus:**

-   Phase 1: Labor surplus is small, driven by the logistic trend near zero plus random noise $\epsilon$. This represents the early **reserve army of labor** described by Marx — present but not yet structurally significant. 

-   Phase 2: New sectors emerge, temporarily increasing labor demand and absorbing surplus workers, which keeps the wage gap low. This is represented by the dampened oscillation: $A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P}$ , when $sin\frac{2t\pi}{P}$ is negative: the oscillation pulls labor surplus below the trend.

-   Phase 3: Labor surplus increases during the accumulation phase. During this phase, new technologies emerges and the capital grows. This leads to overproduction: more products in the market, the price falls, the profits collapse and workers are displaced by new technologies, swelling the reserve army. This is captured by $A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P}$ , when $sin\frac{2t\pi}{P}$ is positive: the oscillation pushes above the trend. With each peak higher than the last as the logistic trend rises underneath.

**Phase 2 and 3 are repeated over T**, each cycle leaves more residual labor surplus than the previous, then **the cumulative reserve army grows (logistic trend rises)**, the growing labor surplus feeds to the growth rate of the wage gap equation $r_w = r_{base,w} + \alpha(labor\_surplus_{t-1})$, and the **wage gap logistic curve steepens until it reaches its inflection point** $t_{0,w}$.

A recovery occurs during the transition from Phase 3 to Phase 2, prompted by the emergence of new sectors and a corresponding expansion of the labor supply. Nevertheless, successive cycles (Phase 2, Phase 3, and recovery) exhibit **heterogeneity in their amplitude and periodicity**. This variance stems fundamentally from the **anarchic movement of the capital.**

#### Estimation of the Autocorrelation Structure

The cumulative pressure series $CP_l$ and $CP_w$ are stochastic. Since each society draws its own sequence of random shocks $\epsilon$ and oscillation realizations, the autocorrelation function estimated from a single society reflects idiosyncratic noise rather than the underlying process structure shared by all societies. To obtain a more stable and representative estimate of the autocorrelation structure, we average across N societies. 
The estimation is further restricted to the post-inflection phase, starting from $t_{0,l}$ and $t_{0,w}$. In the pre-inflection phase, the logistic trend remains close to zero and the cumulative pressures have not yet fully developed, which would bias the autocorrelation estimate downward. The post-inflection phase, by contrast, captures the mature dynamics of the system where structural pressures have fully accumulated and the ARMA autocorrelation structure is most clearly expressed empirically.

**Labor surplus**
 
```{r}
lag_max <- 20 
acf_matrix_l <- matrix(NA, nrow = lag_max , ncol = N_pilot)
acf_matrix_w <- matrix(NA, nrow = lag_max , ncol = N_pilot)
```

```{r}
acf_func <- function(lag_max, acf_matrix, N, T, t0, CP) {
  for (s in 1:N){
    cps <- CP[t0:T, s]
    acf_matrix[,s] <- acf(cps, 
                          lag.max = lag_max,
                          plot = FALSE)$acf[-1]
  }
  acf_avg <- rowMeans(acf_matrix)
  return(acf_avg)
}
```

```{r}
acf_avg_l <- acf_func(lag_max, acf_matrix_l, N_pilot, T_pilot, t0_l, CP_l)
acf_avg_w <- acf_func(lag_max, acf_matrix_w, N_pilot, T_pilot, t0_w, CP_w)
```


```{r}
acf_plot <- function(acf, T, title = "Average ACF"){
  
   plot(1:lag_max, acf,
         type = "h",
         xlab = "Lag",
         ylab = "Average ACF",
         main = title,
         ylim = c(-0.2, 1))
    abline(h = 0)
          # Significance bounds
    abline(h =  1.96 / sqrt(T), lty = 2, col = "#E24B4A")
    abline(h = -1.96 / sqrt(T), lty = 2, col = "#E24B4A")
        
    legend("topright",
            legend = c("Average ACF", "95% significance bounds"),
            col    = c("black", "#E24B4A"),
            lty    = c(1, 2))
  
}
```

```{r}
par(mfrow = c(1,2))
acf_plot(acf_avg_l, T_pilot, title = "ACF labor surplus")
acf_plot(acf_avg_w, T_pilot, title = "ACF wage gap" )

par(mfrow = c(1,1))
```

Both variables exhibit **high autocorrelation as a consequence of their ARMA(1,q) structure**. Yet, unlike the wage gap, which follows a smoother trajectory, the labor surplus features dampened oscillations that cause $CP\_L$ to retain more autocorrelation than $CP\_W$.

#### Derive T from the effective sample size (ESS)

In time series data, observations are not independent — each period carries information from previous ones. The effective sample size T_eff allows us to measure the **true amount of independent information** in the raw sample T when observations are correlated.

$$
T_{eff} = \frac{T}{1+2\sum^{\infty}_{k=1}\rho(k)}
$$


-   $T$ is T_pilot, the raw sample size, the total number of observed periods.
-   $T_{eff}$ is the ESS — the equivalent number of independent observations after accounting for autocorrelation
-   $\rho$ is the autocorrelation structure. 

```{r}
# Labor surplus
Tl_obs <- T_pilot - round(t0_l)
Tl_eff <- Tl_obs/(1+2*sum(acf_avg_l))

# Wage gap
Tw_obs <- T_pilot - round(t0_w)
Tw_eff <- Tw_obs/(1+2*sum(acf_avg_w))

ratio_w <- Tw_eff/Tw_obs
ratio_l <- Tl_eff/Tl_obs

cat("Effective sample size — labor surplus:",
    round(Tl_eff, 3), "\n")
cat("Effective sample size — wage gap:",
    round(Tw_eff, 3), "\n")
cat("Efficiency ratio — wage gap:", 
    round(ratio_w, 3), "\n")
cat("Efficiency ratio — labor surplus:", 
    round(ratio_l, 3), "\n")
cat("Usable decades — labor surplus:", Tl_obs, "\n")
cat("Usable decades — wage gap:", Tw_obs)
```

The effective sample size represents the number of **independent, non-redundant and unique** observations contained in a correlated time series. $Tl_{eff}$ < $Tw_{eff}$ is theorically expected, as the oscillatory component introduces cyclical dependence that reduces the effective information content of $CP\_L$ relative to $CP\_W$. 

The usable decades represent the number of observed periods required to accumulate a given number of truly independent observations. Due to the high autocorrelation structure of each variable, the raw sample is substantially larger than the effective one. For instance, each society must be observed for roughly **150 decades to yield only 5 independent observations from the labor surplus**. Similarly, roughly **140 decades of observation per society are needed to accumulate 15 independent observations from the wage gap.** 

#### Derive N from the events per variable (EPV)

$$
EPV = N\cdot T_{eff}\cdot p
$$

-   $EPV$: expected crisis events
-   $N$: number of societies
-   $T_{eff}$: effective sample size
-   $p$: crisis probability 

Every predictor requires a sufficient number of independent observations to ensure reliable parameter estimates. By setting $T_{eff}$ based on $CP\_W$, we guarantee at least 5 independent observations for  $CP\_L$. 

Recall $P(crisis[t]) = logistic(z)$, with

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$

To estimate N, we vary the crisis probability p across a theoretically motivated range. p represents the baseline crisis probability when $CP\_W = CP\_L = 0$, interpreted as the **minimum crisis probability** at the onset of capitalism before structural contradictions accumulate. 
Setting $CP_W = CP_L = 0$ isolates $\beta_0$ as the sole determinant of the baseline probability, ensuring that the sensitivity analysis varies only the intercept while holding the pressure dynamics fixed.

$$
p (crisis|CP\_W = 0, CP\_L = 0) = \frac{1}{1+e^{-\beta_0}}
$$

A sensitivity analysis is performed on p, deriving the corresponding $\beta_0$ for each value through. It is performed across the range p $\in$ {0.001, 0.005, 0.01, 0.02, 0.05}, reflecting theoretically plausible baseline rates from near-impossible to rare.

$$
\beta_0 = log(\frac{p}{1-p})
$$


```{r}
p <- c(0.001, 0.005, 0.01, 0.02, 0.05)
beta0_values <- qlogis(p)
epv <- 30
T_eff_binding <- Tw_eff

# N calculation for each p
sensitivity_analysis <- data.frame(
  crisis_prob = p ,
  beta0 = round(beta0_values, 4), 
  N_needed = ceiling(epv/(T_eff_binding*p))
)

print(sensitivity_analysis)
```
 
According to the EPV rule, a minimum of 30 crisis events is required across the entire dataset to reliably estimate the three parameters of the logistic model. Since the effective sample size is fixed at $Tw_{eff}$, the number of societies N required to accumulate 30 effective crisis observations is inversely proportional to the baseline crisis probability p. Consequently, lower values of p demand substantially larger N to compensate the rarity of crisis events. 

#### Generate the structural crisis (y_pilot)

The baseline crisis probability p is calibrated from **Turchin and Nefedov's (2009)** observation that recurrent waves of state breakdown occured approximately **three times over five centuries** of European history — the calamitous fourteenth century, the iron century of 1550-1660, and the age of revolutions of 1789-1849. This implies a **per-decade crisis probability of approximately p = 0.006**. This motivates the lower end of our sensitivity analysis, consistent with Marx's argument in Wage Labour and Capital that structural crisis require prolonged accumulation of contradictions before becoming probable. 
The combination that mostly fits with this argument is ($p = 0.005$, $\beta_0 = -5.2933$, $N = 389$). 

Then, we need to determine the parameters $\beta_1$ , $\beta_2$ , $\beta_3$. 

```{r}
beta0 <- -5.2933	
beta1 <- 1.0   # wage gap effect — moderate
beta2 <- 0.5   # labor surplus effect — smaller alone
beta3 <- 2.0   # interaction — largest, captures Marx's threshold
```

The ordering $\beta_3$ > $\beta_1$  > $\beta_2$ reflects three theoretical priorities. The dominance of $\beta_3$ formalizes Marx's conjunctural argument that **structural crisis emerges primarly from the simultaneous occurrence of wage depression and labor surplus.** The ordering beta1 > beta2 reflects the relatively stronger direct effect of wage gap compared to labor surplus in isolation, consistent with Marx's emphasis on wage depression as the most visible manifestation of capitalist contradiction in Wage Labour and Capital.

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$


```{r}
y_func <- function(beta0, beta1, beta2, beta3, CP_l, CP_w, T, N) {
  
  z <- beta0 + (beta1*CP_w) + (beta2*CP_l) + (beta3*CP_l*CP_w)
  p_crisis <- 1/(1+exp(-z))
  
  y <- matrix(rbinom(T*N, size = 1, prob = p_crisis), 
              ncol = N, nrow = T)
  
  return(y)
}
```

```{r}
y_pilot <- y_func(beta0, beta1, beta2, beta3, CP_l, CP_w, T_pilot, N_pilot)
```

```{r}
head(y_pilot)
```

#### Data Generating Process : Final Simulation

For the final simulation, we set **N_final = 389 societies and T_final = 200**. This timeline spans the necessary pre- and post-inflection decades. The choice to include exactly 150 post-inflection decades is driven by the differing autocorrelation structures of our variables: while these 150 decades translate to merely 5 effective independent informations for labor surplus, they yield 15 independent observations for the wage gap. The pre-inflection phase is retained to preserve the theoretical warmup consistent with Marx's early capitalism argument. However, the labor surplus exhibits high autocorrelation, substantially reducing the effective independent information per society. **To compensate for this loss of independence and to satisfy the EPV rule of a minimum of 30 effective crisis, N = 389 societies are required.**


```{r}
N_final <- 389
T_final <- 200
```

```{r}
# Generate labor surplus , wage gap, CP_l, CP_w, y with N_final, T_final 
labor_surplus_fin <- func_labor_surplus(N_final, T_final)
wage_gap_fin <- func_wage_gap(N_final, T_final, labor_surplus_fin)

cp_fin <- cp(N_final, T_final, labor_surplus_fin, wage_gap_fin)
CPfin_l <- cp_fin$CP_l
CPfin_w <- cp_fin$CP_w

y_final <- y_func(beta0, beta1, beta2, beta3, CPfin_l, CPfin_w, T_final, N_final)
```


```{r}
# crisis event count comparison
data.frame (
  Simulation = c("Pilot","Final"),
  N = c(N_pilot, N_final),
  T = c(T_pilot, T_final),
  Total_crisis = c(sum(y_pilot), sum(y_final))
    )
```

The substantial jump from 1328 to 51356 crises is an expected outcome rather than a anomaly; it directly reflects the roughly 39-fold scaling of the number of societies from $N_{pilot} = 10$ to $N_{final} = 389$

#### Parameter recovery analysis

The true parameters are known by construction. By fitting the logistic regression, we can validate whether the sample size N derived from the EPV rule is sufficient to recover the true parameters, assess if a standard logistic regression remains reliable under the ARMA(1,q) autocorrelation structure of $CP_L$ and $CP_W$ and test whether the parameter $\beta_3$ is detectable. Here, $\beta_3$ represents the mathematical formalization of Marx's argument that the interaction of wage depression and labor surplus amplifies crisis emergence. This approach allows us to determine whether the statistical model can recover known parameters from synthetic data generated under controlled conditions. **If the estimated parameters diverge substantially from the true parameters, the model certainly cannot be trusted to estimate unknown parameters from real historical data**

To fit the logistic regression, we will utilize two estimation approaches and evaluate the differences in the resulting coefficients :

**1. Long format + glm()**

This is the cleanest approach for logistic regression in R: the NxT matrices to a long format dataframe and fit the data straightforwardly. 

```{r}
df_long <- data.frame(
  society = rep(1:N_final, each = T_final), 
  decade = rep(1:T_final, times = N_final), 
  CP_w = as.vector(CPfin_w), 
  CP_l = as.vector(CPfin_l), 
  y = as.vector(y_final)
)
```

```{r}
model <- glm(y~CP_w+CP_l+CP_w:CP_l, 
             data = df_long, 
             family = binomial(link = "logit"))
summary(model)
```

**2. Custom Maximum Likelihood Estimation function**

This method shows what glm() is doing internally.

```{r}
X <- cbind(
  intercept = 1, 
  CP_w = as.vector(CPfin_w), 
  CP_l = as.vector(CPfin_l), 
  interaction = as.vector(CPfin_w)*as.vector(CPfin_l)
)

y_vec <- as.vector(y_final)
```



```{r}
log_likelihood <- function(beta, X, y) {
  
  z <- X %*% beta
  p <- 1/(1+exp(-z))
  ll <- sum(y*log(p + 1e-10) + (1-y)*log(1-p  + 1e-10))
  
  return(-ll)
}
```

```{r}
results <- optim(
    par = c(0,0,0,0),
    fn = log_likelihood,
    X = X, 
    y = y_vec, 
    method = "BFGS", 
    hessian = TRUE
  )
```


```{r}
data.frame(
  true = c(beta0, beta1, beta2, beta3), 
  estimated_1 = c(coef(model)), 
  estimated_2 = c(results$par)
)
```

The close alignment between the estimated and true parameters demonstrates that both models successfully recover the parameters derived from Marxian theory. This also validates our use of the EPV rule, confirming that the sample size was sufficient for reliable estimation. Furthermore, the estimated parameters maintain the expected ordering ($\beta_3$ > $\beta_2$ > $\beta_1$), providing empirical support for the theoretical argument regarding the dominance of the interaction term. Finally, despite the autocorrelated structure of the CP variables, standard logistic regression via glm() performs comparably to the custom MLE. This suggests that, under this specific simulation design, the autocorrelation does not introduce severe bias into the estimates.

#### Time Series Diagnosis :

Assumption violation > Coefficient biased > overconfident model > inaccurate standard errors

**1.    Comparison of the ACF of the pilot and final simulation :**

```{r}
acfin_matrix_l <- matrix(NA, nrow = lag_max , ncol = N_final)
acfin_matrix_w <- matrix(NA, nrow = lag_max , ncol = N_final)

acfin_avg_l <- acf_func(lag_max, acfin_matrix_l, N_final, T_final, t0_l, CPfin_l)
acfin_avg_w <- acf_func(lag_max, acfin_matrix_w, N_final, T_final, t0_w, CPfin_w)
```

```{r}
# compare acf pilot and acf final for labor surplus
acf_comp_l<- data.frame(lag = 1:lag_max, 
                        acf_final = acfin_avg_l, 
                        acf_pilot = acf_avg_l)

# compare acf pilot and acf final for wage gap
acf_comp_w <- data.frame(lag = 1:lag_max, 
                         acf_final = acfin_avg_w, 
                         acf_pilot = acf_avg_w)
```

```{r}
acf_comparison <- function(acf_comp, title = "ACF Comparison: Final vs Pilot"){
  
  # convert to long format 
  acf_long <- pivot_longer(acf_comp, 
                           cols = c(acf_final, acf_pilot), 
                           names_to = "Type", 
                           values_to = "ACF_Values")
  
  p <- ggplot(acf_long, aes(x = lag, y = ACF_Values, color = Type, group = Type)) +
    geom_line(position = position_dodge(width = 0.2), linewidth = 1) + 
    geom_point(position = position_dodge(width = 0.2), size = 3) + 
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
    theme_minimal() +
   
    labs(title = title,        
         x = "Lag", 
         y = "ACF Value") +
    scale_color_manual(values = c("acf_final" = "steelblue", "acf_pilot" = "firebrick"))
  
  # Return the plot
  return(p) 
}
```

```{r}
plot_wage <- acf_comparison(acf_comp_w, title = "ACF Comparison (wage gap)" )
plot_labor <-acf_comparison(acf_comp_l, title = "ACF Comparison (labor surplus)" )
```

```{r}
combined_plot <- plot_wage + plot_labor
combined_plot
```

As confirmed by the plots, the final and pilot simulations exhibit the **same autocorrelation structure.** This alignment indicate two key points: 

-   **Stability of the Data Generating Process (DGP):** The DGP is stable, meaning the underlying parameters consistently produce the same autocorrelation structure regardless of the sample size (N)

-   **Validity of the Effective Sample Size (ESS):** The ESS derived from the pilot simulation remains valid for the final simulation, as the underlying autocorrelation structure has remained unchanged. 

**2.   Residual autocorrelation from logistic regression**

The ARMA(1,q) structure of $CP_w$ and $CP_l$ violates the logistic regression assumption of independent observations. However, this does not necessarily bias the estimated $\beta$ coefficients; rather, the primary consequence is **inaccurate standard errors**. Because $CP_w$ and $CP_l$ are expected to capture the temporal structure, the residuals should be uncorrelated. **The presence of correlated residuals can lead to biased standard errors.**

```{r}
residuals_pearson <- residuals(model, type = "pearson")
lag_max <- min(Tl_obs/4, 20)
df_long_ord <- df_long %>% arrange(society, decade)
acf_resid_matrix <- matrix (NA, nrow = lag_max, ncol = N_final)


for (s in 1:N_final){
  # Extract the residuals for society s
  res_s <- residuals_pearson[df_long$society == s]
  acf_resid_matrix[,s]<-acf(res_s, lag_max, plot = FALSE)$acf[-1]
}

avg_resid_matrix <- rowMeans(acf_resid_matrix)
```

```{r}
acf_plot(avg_resid_matrix, T_final, title = "ACF residuals")
```

Because the predictors $CP_W$ and $CP_L$ effectively explain the autoregressive nature of the data, the model residuals exhibit negligible autocorrelation, as demonstrated in the plots below.

#### Bayesian Version :

The previous simulations used a frequentist framework that assumes parameters are fixed values. However, because the goal of this project is to model the challenges of historical data — inspired by Marx's Wage Labour and Capital, we use synthetic data that mimics the non-independence of historical data with the ARMA(1,q) structure of the observations. Because the data points are not truly independent, the Bayesian approach provides a more honest and accurate quantification of uncertainty.

*Note on Limitations:* While this simulation successfully replicates the temporal autocorrelation typical of historical data, it assumes the dataset is fully complete. **Real historical archives are often fragmented and incomplete.** Future iterations of this model should introduce missingness mechanisms to fully mimic the challenges of empirical historical research.

**1.Manual Metropolis Hasting** :

For our manual Metropolis-Hastings implementation, we assign Normal priors to the model parameters. Specifically, we center the priors for the parameters $\beta_0$, $\beta_1$, $\beta_2$ and $\beta_3$ at zero to reflect a neutral initial assumption.

```{r}
# log prior function 
log_prior <- function(beta){
  # beta0~Normal(0, 10)
  # beta1, beta2, beta3~Normal(0, 5)
  dnorm(beta[1], 0, 10, log = TRUE)+ 
  dnorm(beta[2], 0, 5, log = TRUE)+ 
  dnorm(beta[3], 0, 5, log = TRUE)+ 
  dnorm(beta[4], 0, 5, log = TRUE)
}

# Same log_likehood as we use in the manual implementation of the logistic regression.
log_posterior <- function(beta, X, y){
  log_prior(beta) + log_likelihood(beta, X, y)
}
```

```{r}
manual_metropolis_hasting <- function(X, y, n_iter = 5000, proposal_sd = .05, model){
  
  # initialize the storage
  samples <- matrix(NA, nrow = n_iter, ncol = 4)
  colnames(samples) <- c("beta0","beta1","beta2", "beta3")
  
  # starting point 
  beta_current <- coef(model)
  
  # initialize the acceptance
  acceptance <- 0
  
  for (i in 1:n_iter){
    # add random noise to beta_current to get the beta_proposed
    beta_proposed <- beta_current + rnorm(4, mean = 0, sd = proposal_sd)
    # compute the log posterior ratio
    log_ratio <- log_posterior(beta_proposed, X, y) - log_posterior(beta_current, X, y)
    # accept or reject 
    if (log(runif(1)) < log_ratio) {
      beta_current <- beta_proposed # update beta_current
      acceptance <- acceptance + 1
    }
    samples[i,] <- beta_current
  }
  cat("Acceptance rate: ", round(acceptance/n_iter, 3), "\n")
  
  return(samples)
}
```

```{r}
set.seed(42)
mcmc_samples <- manual_metropolis_hasting(X, y_vec, n_iter = 5000, proposal_sd = .05, model)
```

```{r}
trace_plot <- function(sample){

    for (j in 1:4) {
      plot(sample[, j],
           type = "l",
           main = paste("Trace —", colnames(sample)[j]),
           xlab = "Iteration",
           ylab = "Value",
           col  = "#378ADD")
      abline(h = c(beta0, beta1, beta2, beta3)[j],
             col = "#E24B4A", lwd = 2, lty = 2)
    }
}

post_plot <- function(samples){
    # discard warm up 
  warm_up <- 1000 
  post_samples <- samples[(warm_up+1):(nrow(samples)), ]
  

  for(j in 1:4){
    param_name <- colnames(post_samples)[j]
    true_value <- c(beta0, beta1, beta2, beta3)
    
    hist(post_samples[,j], 
         breaks = 50, 
         main   = paste("Posterior of", param_name),
         xlab   = param_name,
         col    = "#378ADD",
         border = "white",
         freq   = FALSE)
    
      # true value
    abline(v = true_value, col = "#E24B4A", lwd = 2, lty = 2)
    
    # posterior mean
    abline(v = mean(post_samples[, j]), 
           col = "#1D9E75", lwd = 2)
    
    legend("topright",
           legend = c("True value", "Posterior mean"),
           col    = c("#E24B4A", "#1D9E75"),
           lty    = c(2, 1), lwd = 2, cex = 0.7)
  }
}
```


```{r}
par(mfcol = c(2, 4), mar = c(4, 4, 2, 1))
trace_plot(mcmc_samples)
post_plot(mcmc_samples)
par(mfrow = c(1, 1))
```
**With Standardized observations:** To optimize the Metropolis-Hastings algorithm, the predictor matrix was standardized. Putting all variables on the same scale prevents numerical instability and symmetrizes the likelihood surface, making it much easier for the algorithm's random walk to navigate the parameters efficiently.

```{r}
# standardize X
X_std <- cbind(
  intercept = 1, 
  CP_w = scale(as.vector(CPfin_w))[,1], 
  CP_l = scale(as.vector(CPfin_l))[,1], 
  interaction = scale(as.vector(CPfin_w)*as.vector(CPfin_l))[,1]
)
```

```{r}
model_std <- glm(y_vec~X_std-1, family = binomial)
```

```{r}
set.seed(42)
mcmc_std <- manual_metropolis_hasting(X_std, y_vec, n_iter = 5000, proposal_sd = .05, model_std)
```

```{r}
par(mfcol = c(2, 4), mar = c(4, 4, 2, 1))
trace_plot(mcmc_std)
post_plot(mcmc_std)
par(mfrow = c(1, 1))
```

The correlation between $CP_W$ and $CP_L$ , exacerbated by their interaction term, results in **extreme multicollinearity.** The observations do not contain enough independent variation to precisely estimate all four parameters.
Consequently, a standard Metropolis-Hastings algorithm may struggle to converge to the true parameter values. By implementing a **multivariate Normal proposal**, the algorithm can learn and navigate this underlying covariance structure.


```{r}
multivariate_normal_proposal <- function(X, y, n_iter = 5000, proposal_sd = .05, model){
  
  # initialize the storage
  samples <- matrix(NA, nrow = n_iter, ncol = 4)
  colnames(samples) <- c("beta0","beta1","beta2", "beta3")
  
  # starting point 
  beta_current <- coef(model)
  
  # initialize the acceptance
  acceptance <- 0
  
  step_cov <- vcov(model)*proposal_sd^2
  
  for (i in 1:n_iter){
    # add random noise to beta_current to get the beta_proposed
    beta_proposed <- mvrnorm(n = 1, mu = beta_current, Sigma = step_cov)
    # compute the log posterior ratio
    log_ratio <- log_posterior(beta_proposed, X, y) - log_posterior(beta_current, X, y)
    # accept or reject 
    if (log(runif(1)) < log_ratio) {
      beta_current <- beta_proposed # update beta_current
      acceptance <- acceptance + 1
    }
    samples[i,] <- beta_current
  }
  cat("Acceptance rate: ", round(acceptance/n_iter, 3), "\n")
  
  return(samples)
}
```

```{r}
set.seed(42)
mnp_samples <- multivariate_normal_proposal(X, y_vec, n_iter = 5000, proposal_sd = .05, model)
```

```{r}
par(mfcol = c(2, 4), mar = c(4, 4, 2, 1))
trace_plot(mnp_samples)
post_plot(mnp_samples)
par(mfrow = c(1, 1))
```

```{r}
mnp_std <- multivariate_normal_proposal(X_std, y_vec, n_iter = 5000, proposal_sd = .05, model_std)
```

```{r}
par(mfcol = c(2, 4), mar = c(4, 4, 2, 1))
trace_plot(mnp_std)
post_plot(mnp_std)
par(mfrow = c(1, 1))
```

##### R-hat computation for the Manual Metropolis Hasting :

$\hat{R}$ is a convergence diagnostic. It tells whether the MCMC chains have converged to the same posterior distribution.

**Between-chain variance** : This measures how different the two chain means are from each other.

$$
B = \frac{N}{M-1} \sum^M_{m=1} (\hat{\psi}_m - \hat{\psi})^2
$$

**Within-chain variance** : This measures how much each chain varies internally around its own mean.

$$
W = \frac{1}{M} \sum^M_{m=1} s^2_m
$$

**Estimated marginal posterior variance** : This combines B and W into an estimate of the true posterior variance.

$$
\hat{var} = \frac{N-1}{N}W + \frac{1}{N}B
$$

**R-hat** : 

$$
R = \sqrt \frac{\hat{var}}{W}
$$

```{r}
compute_rhat <- function(samples) {
  
  n_params <- ncol(samples) 
  rhat <- numeric(n_params) # create an empty vector of Rhat
  
  for (j in 1:n_params) {
    chain <- samples[,j]
    N_total <- length(chain)
    
    # split the chain
    half <- floor(N_total/2)
    chain1 <- chain[1:half]
    chain2 <- chain[(half+1):(half*2)] # to make sure chain1 and chain2 have the same length
    
    # set N and M
    N <- half # length of each split chain
    M <- 2 # number of chains
    
    # chains mean
    psi_bar_1 <- mean(chain1)
    psi_bar_2 <- mean(chain2)
    psi <- mean(psi_bar_1, psi_bar_2)
    
    # between-chain variance
    B <- (N/(M-1)) * sum( (psi_bar_1 - psi)^2, (psi_bar_2 - psi)^2)
    
    # within chain variance 
    W <- (var(chain1)+var(chain2))/M
    
    # estimated marginal posterior variance 
    var_hat <- ((N-1)/N)* W + (1/N)*B
    
    # R-hat 
    rhat[j] <- sqrt(var_hat/W)
  }
  return(rhat)
}
```

```{r}
rhat_mcmc_samples <- compute_rhat(mcmc_samples)
rhat_mcmc_std <- compute_rhat(mcmc_std)
rhat_mnp_samples <- compute_rhat(mnp_samples)
rhat_mnp_std <- compute_rhat(mnp_std)
```

```{r}
data.frame(
  parameters = c("beta0", "beta1", "beta2", "beta3"), 
  rhat1 = c(rhat_mcmc_samples), 
  rhat2 = c(rhat_mcmc_std), 
  rhat3 = c(rhat_mnp_samples), 
  rhat4 = c(rhat_mnp_std)
)
```

```{r, results='hide'}
bayesian_model <- brm(
  formula = y~CP_w+CP_l+CP_w:CP_l, 
  data = df_long, 
  family = bernoulli(link = logit), 
  prior = c(
    prior(normal(0, 10), class = "Intercept"), 
    prior(normal(0, 5), class = "b")
  ), 
  chain = 4, 
  iter = 2000, 
  warmup = 1000, 
  seed = 42, 
  cores = 4, 
  control = list(
    adapt_delta = .95, 
    max_treedepth = 12
  )
)
```

```{r}
plot(bayesian_model)
```

```{r}
summary_bayesian <- summary(bayesian_model)
rhat <- summary_bayesian$fixed[, "Rhat"]
cat("R-hat values for HMC:, \n")
print(round(rhat, 4))
```






