---
title: "Wage Labour and Capital"
output: github_document
---

In this project, we translated the theoretical framework of Marx's Wage Labour and Capital into a mathematical model.
To ensure that the variables behave according to the theoretical mechanisms describes in the literature, rather than being confounded by the complexity of real historical processes, we rely on synthetic data rather than empirical historical data. This allow us to control the data generating process precisely. 

This simulation is structured around two independent variables : **wage gap** and **labor surplus** whose dynamics are governed by theoretically motivated equations developed in the following sections. These variables feed into **cumulative pressure terms** ($CP_w$, $CP_l$), which in turn predict the probability of **structural crisis** through a logistic regression model.

We observed N societies over T decades. To determine the sample size, we used the 10 events per variable (EPV) rule of thumb for logistic regression. But two methodological challenges must be addressed. First, the determination of N and T depends on the **crisis probability and the autocorrelation structure of the cumulative pressure variables.** Second, the cumulative pressure terms exhibit an **ARMA (1,q)** autocorrelation structure, making analytical derivation of the effective sample size (ESS) inappropriate. For these reasons, we proceed in two stages: a pilot simulation with provisional parameters to empirically **estimate the autocorrelation structure**, followed by the full simulation with final **N and T derived from both the empirical autocorrelation estimate and the sensitivity analysis on crisis probability.**

Natural order : estimation of the empirical autocorrelation from the pilot, derive T with the effective sample size (ESS) and N with the events per variable (EPV).

#### The Variables : 

A structural crisis occurrence at decade t can be translated in a logistic regression with the possible output : *crisis, no crisis.*

$P(crisis[t]) = logistic(z)$, with :

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$

- $x_1$: The wage gap measures the distance between the wage and the subsistence cost.

- $x_2$: The labor surplus is the excess of workers relative to available job

- $x_1x_2$ is the interaction term. It captures the idea that the effect of one variable depends on the level of the others. So when wage gap and labor surplus are high simultaneously, the combine effect on crisis probability is greater than their individual effects added together. The interaction term formalizes Marx's argument that structural crisis emerges not from wage depression or labor surplus independently, but from their simultaneous occurrence — consistent with the threshold mechanism described in Wage Labour and Capital.

Following the **Events Per Variable rule**, which recommends a minimum of 10 observed events per predictor to ensure reliable parameter estimation in logistic regression, and given that our model contains 3 variables ($x_1$, $x_2$, $x_1x_2$), the minimum number of **expected crisis events is 3×10=30**

#### Data Generating Process 

```{r}
set.seed(123)
```


```{r}
N_pilot <- 10
T_pilot <- 200
```

```{r}
# labor_surplus parameters
K_l <- 1
r_l <- .05
t0_l  <- 50
A <- .2
P <- 5
delta_l <- .43
```

```{r}
# wage_gap parameters
K_w <- 1
r_base <- .03
alpha <- .5
t0_w <- 60
```

```{r}
# decay parameter
delta_c <- 0.43
```

```{r}
# Initialize storage
labor_surplus <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
wage_gap <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
CP_l <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
CP_w <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
```


#### Generate wage gap and labor surplus
Labor surplus has to be generated before wage gap because the growth rate $r_w$ of the logistic trend of the wage gap depends on labor surplus

1.    Labor surplus : it represents the excess of workers relative to available job

$$
labor\_surplus (t) = \frac{K_l}{1+e^{-r_l(t-t_{0,l})}} + A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P} + \epsilon
$$

```{r}
for (n in 1:N_pilot) {
  for (t in 1:T_pilot) {
    
    logistic_trend <- K_l/(1+exp(-r_l*(t-t0_l)))
    oscillation <- A * exp(-delta_l*t) * sin((2*t*pi)/P)
    epsilon <- rnorm(1, mean = 0, sd = .02)
    
    labor_surplus[t, n] <- logistic_trend + oscillation + epsilon
  }
}
```

```{r}
head(labor_surplus)
```


2.    Wage gap : it measures the distance between the wage and the subsistence cost.


$$
wage\_gap(t) = \frac{K_w}{1+e^{-r_w(t-t_{0,w})}}
$$

$$
r_w = r_{base,w} + \alpha(labor\_surplus_{t-1})
$$

```{r}
for (n in 1:N_pilot) {
  for (t in 1:T_pilot) {
    
    if (t == 1) {r_w <- r_base} else {r_w <- r_base + alpha*(labor_surplus[t-1, n])}
    
    wage_gap[t,n]<- K_w/(1+exp(-r_w*(t-t0_w)))
  }
}
```

```{r}
head(wage_gap)
```

The logistic formula of the wage gap is deterministic given the growth rate $r_w$ and the time $t$. 
At $t=1$, $r_w = r_{base}$ because there is no labor surplus before $t=1$. Therefore, the wage gap at $t=1$ is identical for all societies, meaning that all societies start with the same structural conditions but diverge as the market history evolves. 

#### Generate cumulative wage gap and cumulative labor surplus
Weights are assigned inversely to chronological distance, prioritizing recent events over earlier ones via an exponential decay factor.

$$
CP\_L(t) = \sum^{t-1}_{k} L_{t-k}\cdot e^{-\delta_c\cdot k}
$$

$$
CP\_W(t) = \sum^{t-1}_{k} W_{t-k}\cdot e^{-\delta_c\cdot k}
$$

```{r}
for (n in 1:N_pilot){
  for (t in 1:T_pilot){
    
    # cumulative pressure is only meaningful at t=2
    if (t==1){
      CP_l[t, n] <- 0
      CP_w[t, n] <- 0
    }
    else {
      
      cpsum_l <- 0
      cpsum_w <- 0
      
      for (k in 1:(t-1)){
        cpsum_l <- cpsum_l + labor_surplus[t-k, n] * exp(-delta_c*k)
        cpsum_w <- cpsum_w + wage_gap[t-k, n] * exp(-delta_c*k)
      }
      
      CP_l[t, n] <- cpsum_l
      CP_w[t, n] <- cpsum_w
    }
  }
  }
   
```

```{r}
head(CP_l)
```

```{r}
head(CP_w)
```

-   The identical wage gap at $t=2$ across all societies can be explain by the identical wage gap at $t=1$. At $t=2$, CP_w looks back one period, meaning it only uses data in $t=1$.

```{r}
plot(1:T_pilot, CP_l[,1], type = "l", 
     xlab = "decades", ylab = "Cumulative pressure",
     main = "cumulative labor surplus and wage gap for society 1")
lines(1:T_pilot, CP_w[,1],col ="red")
legend("topleft", legend = c("CP_l", "CP_w"), col = c("black", "red"), lty = 1)
```

**Phases of the labor surplus:**

"Wages will now rise, now fall,according to the relation of supply and demand, according as competition shapes itself between the buyers of labour-power, the capitalists, and the sellers of labour-power, the workers. "

-   *Phase 1: * Labor surplus is small and pre-inflection, driven by the logistic trend near zero plus random noise $\epsilon$. This represents the early **reserve army of labor** described by Marx — present but not yet structurally significant. 

-   *Phase 2: * New sectors emerge, temporarily increasing labor demand and absorbing surplus workers, which keeps the wage gap low. This is represented by the dampened oscillation: $A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P}$ , when $sin\frac{2t\pi}{P}$ is negative: the oscillation pulls labor surplus below the trend.

-   *Phase 3: * Labor surplus increases during the accumulation phase. During this phase, new technologies emerges and the capital grows. This leads to overproduction: more products in the market, the price falls, the profits collapse and profits collapse and workers are displaced by new technologies, swelling the reserve army. This is captured by $A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P}$ , when $sin\frac{2t\pi}{P}$ is positive: the oscillation pushes above the trend. With each peak higher than the last as the logistic trend rises underneath.

```{r}
n <- 40
trend_l <- K_l / (1 + exp(-r_l * (10:n - t0_l)))

# Main plot
plot(10:n, labor_surplus[10:n, 1], type = "n",
     xlab = "Decades", ylab = "Labor Surplus",
     main = "Labour Surplus with Logistic Trend - Society 1")


# Add lines on top of shading
lines(10:n, labor_surplus[10:n, 1], col = "#00070d", lwd = 1.5)
lines(10:n, trend_l, col = "#E24B4A", lty = 2, lwd = 2)

# Legend
legend("topleft",
       legend = c("Labor surplus", "Logistic trend"),
       col = c("#00070d", "#E24B4A"),
       lty = c(1, 2, 1, 1), lwd = 2)
```

The plot below represent the Phase 2 and Phase 3 of the labor surplus from decade 10 to decade 40. The dotted line is the **logistic trend**: Phase 2 is the labor surplus below the trend, and Phase 3 is the labor surplus above the trend. 

Phase 2 and 3 are repeated over T, each cycle leaves more residual labor surplus than the previous, then the cumulative reserve army grows (logistic trend rises), the growing labor surplus feeds to the growth rate of the wage gap equation $r_w = r_{base,w} + \alpha(labor\_surplus_{t-1})$, and the wage gap logistic curve steepens until it reaches its inflection point $t_{0,w}$.

A recovery occurs during the transition from Phase 3 to Phase 2, prompted by the emergence of new sectors and a corresponding expansion of the labor supply. Nevertheless, successive cycles (Phase 2, Phase 3, and recovery) exhibit heterogeneity in their amplitude and periodicity. This variance stems fundamentally from the **anarchic movement of capital.**

#### Estimation of the Autocorrelation Structure

The cumulative pressure series $CP_l$ and $CP_w$ are stochastic. Since each society draws its own sequence of random shocks $\epsilon$ and oscillation realizations, the autocorrelation function estimated from a single society reflects idiosyncratic noise rather than the underlying process structure shared by all societies. To obtain a more stable and representative estimate of the autocorrelation structure, we average across N societies. 
The estimation is further restricted to the post-inflection phase, starting from $t_{0,l}$ and $t_{0,w}$. In the pre-inflection phase, the logistic trend remains close to zero and the cumulative pressures have not yet fully developed, which would bias the autocorrelation estimate downward. The post-inflection phase, by contrast, captures the mature dynamics of the system where structural pressures have fully accumulated and the ARMA autocorrelation structure is most clearly expressed empirically.

**Labor surplus**
 
```{r}
lag_max_l <- 20 
acf_matrix_l <- matrix(NA, nrow = lag_max_l , ncol = N_pilot)

for (s in 1:N_pilot) {
  cps_l <- CP_l[t0_l:N_pilot, s]
  acf_matrix_l [,s] <- acf(cps_l,
                    lag.max = lag_max_l,
                    plot = FALSE)$acf[-1]  # remove lag-0 (always 1)
}

acf_avg_l <- rowMeans(acf_matrix_l)

```

**Wage gap**

```{r}
lag_max_w <- 20 
acf_matrix_w <- matrix(NA, nrow = lag_max_w , ncol = N_pilot)

for (s in 1:N_pilot) {
  cps_w <- CP_w[t0_w:N_pilot, s]
  acf_matrix_w [,s] <- acf(cps_w,
                    lag.max = lag_max_w,
                    plot = FALSE)$acf[-1]  # remove lag-0 (always 1)
}

acf_avg_w <- rowMeans(acf_matrix_w)

```

#### Derive T from the effective sample size (ESS)

In time series data, observations are not independent — each period carries information from previous ones. The effective sample size T_eff allows us to measure the **true amount of independent information** in the raw sample T when observations are correlated.

$$
T_{eff} = \frac{T}{1+2\sum^{\infty}_{k=1}\rho(k)}
$$


-   $T$ is T_pilot, the raw sample size, the total number of observed periods.
-   $T_{eff}$ is the ESS — the equivalent number of independent observations after accounting for autocorrelation
-   $\rho$ is the autocorrelation structure. 

```{r}
# Labor surplus
Tl_obs <- T_pilot - round(t0_l)
Tl_eff <- Tl_obs/(1+2*sum(acf_avg_l))

# Wage gap
Tw_obs <- T_pilot - round(t0_w)
Tw_eff <- Tw_obs/(1+2*sum(acf_avg_w))

ratio_w <- Tw_eff/Tw_obs
ratio_l <- Tl_eff/Tl_obs

cat("Effective sample size — labor surplus:",
    round(Tl_eff, 3), "\n")
cat("Effective sample size — wage gap:",
    round(Tw_eff, 3), "\n")
cat("Efficiency ratio (T_eff/T_usable) — wage gap:", 
    round(ratio_w, 3), "\n")
cat("Efficiency ratio (T_eff/T_usable) — labor surplus:", 
    round(ratio_l, 3), "\n")
cat("Usable decades for labor surplus (Tl_usable):" ,
    round(Tl_eff/ratio_l, 3), "\n")
cat("Usable decades for wage gap (Tw_usable):" ,
    round(Tw_eff/ratio_w, 3))
```

The effective sample size represents the number of **independent, non-redundant and unique** observations contained in a correlated time series. Given the dampened oscillations in labor surplus, the cumulative pressure $CP\_L$ carries additional autocorrelation than $CP\_W$, which follows a smoother trajectory. Consequently, $Tl_{eff}$ < $Tw_{eff}$ is theorically expected, as the oscillatory component introduces cyclical dependence that reduces the effective information content of $CP\_L$ relative to $CP\_W$. 

The usable decades represent the number of observed periods required to accumulate a given number of truly independent observations. Due to the high autocorrelation structure of each variable, the raw sample is substantially larger than the effective one. For instance, each society must be observed for roughly **150 decades to yield only 11 independent observations from the labor surplus**. Similarly, roughly **140 decades of observation per society are needed to accumulate 20 independent observations from the wage gap.** 

#### Derive N from the events per variable (EPV)

$$
EPV = N\cdot T_{eff}\cdot p
$$

-   $EPV$: expected crisis events
-   $N$: number of societies
-   $T_{eff}$: effective sample size
-   $p$: crisis probability 

Every predictor requires a sufficient number of independent observations to ensure reliable parameter estimates. By setting $T_{eff}$ based on $CP\_L$, we guarantee at least 11 independent observations for the most autocorrelated variable. Meanwhile, $CP\_W$ automatically yields an excess of observations given its lower autocorrelation, making $CP\_L$ the binding constraint.

Recall $P(crisis[t]) = logistic(z)$, with

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$

To estimate N, we vary the crisis probability p across a theoretically motivated range. p represents the baseline crisis probability when $CP\_W = CP\_L = 0$, interpreted as the **minimum crisis probability** at the onset of capitalism before structural contradictions accumulate. 
Setting $CP_W = CP_L = 0$ isolates beta_0 as the sole determinant of the baseline probability, ensuring that the sensitivity analysis varies only the intercept while holding the pressure dynamics fixed.

$$
p (crisis|CP\_W = 0, CP\_L = 0) = \frac{1}{1+e^{-\beta_0}}
$$

A sensitivity analysis is performed on p, deriving the corresponding $\beta_0$ for each value through. It is performed across the range p $\in$ {0.001, 0.005, 0.01, 0.02, 0.05}, reflecting theoretically plausible baseline rates from near-impossible to rare.


$$
\beta_0 = log(\frac{p}{1-p})
$$


```{r}
p <- c(0.001, 0.005, 0.01, 0.02, 0.05)
beta0_values <- qlogis(p)
epv <- 30
T_eff_binding <- Tl_eff

# N calculation for each p
sensitivity_analysis <- data.frame(
  crisis_prob = p ,
  beta0 = round(beta0_values, 4), 
  N_needed = ceiling(epv/(T_eff_binding*p))
)

print(sensitivity_analysis)
```
 
According to the EPV rule, a minimum of 30 crisis events is required across the entire dataset to reliably estimate the three parameters of the logistic model. Since the effective sample size is fixed at $T_{eff}$, the number of societies N required to accumulate 30 effective crisis observations is inversely proportional to the baseline crisis probability p. Consequently, lower values of p demand substantially larger N to compensate the rarity of crisis events. 

#### Generate the structural crisis

The baseline crisis probability p is calibrated from Turchin and Nefedov's (2009) observation that recurrent waves of state breakdown occured approximately **three times over five centuries** of European history — the calamitous fourteenth century, the iron century of 1550-1660, and the age of revolutions of 1789-1849. This implies a **per-decade crisis probability of approximately p = 0.006**. This motivates the lower end of our sensitivity analysis, consistent with Marx's argument in Wage Labour and Capital that structural crisis require prolonged accumulation of contradictions before becoming probable. 
The combination that mostly fits with this argument is ($p = 0.005$, $\beta_0 = -5.2933$, $N = 504$). 

Then, we need to fix the parameters $\beta_1$ , $\beta_2$ , $\beta_3$. 

```{r}
beta0 <- -5.2933	
beta1 <- 1.0   # wage gap effect — moderate
beta2 <- 0.5   # labor surplus effect — smaller alone
beta3 <- 2.0   # interaction — largest, captures Marx's threshold
```

The ordering $\beta_3$ > $\beta_1$  > $\beta_2$ reflects three theoretical priorities. The dominance of $\beta_3$ formalizes Marx's conjunctural argument that structural crisis emerges primarly from the simultaneous occurrence of wage depression and labor surplus. The ordering beta1 > beta2 reflects the relatively stronger direct effect of wage gap compared to labor surplus in isolation, consistent with Marx's emphasis on wage depression as the most visible manifestation of capitalist contradiction in Wage Labour and Capital.

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$

```{r}
z <- beta0 + beta1*CP_w + beta2*CP_l + beta3*CP_l*CP_w
p_crisis <- 1/(1+exp(-z))

y <- matrix(rbinom(T_pilot*N_pilot, size = 1, prob = p_crisis), 
            ncol = N_pilot, nrow = T_pilot)
```

```{r}
head(y)
```

#### Final simulation

For the final simulation, we set **N = 504 societies and T = 200**, corresponding to the inflection point of labor surplus $t_{0,l}$ plus the post-inflection usable window 150 decades. The pre-inflection phase is retained to preserve the theoretical warmup consistent with Marx's early capitalism argument. However, the labor surplus exhibits high autocorrelation, substantially reducing the effective independent information per society. To compensate for this loss of independence and to satisfy the EPV rule of a minimum of 30 effective crisis, N = 504 societies are required. 

```{r}

```






















