---
title: "Wage Labour and Capital"
output: github_document
---

A structural crisis occurrence at decade t can be translated in a logistic regression with the possible output : *crisis, no crisis.*

$P(crisis[t]) = logistic(z)$, with :

$$
z = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2
$$

- $x_1$: wage gap

- $x_2$: labor surplus

- $x_1x_2$ is the interaction term. It captures the idea that the effect of one variable depends on the level of the others. So when wage gap and labor surplus are high simultaneously, the combine effect on crisis probability is greater than their individual effects added together. The interaction term formalizes Marx's argument that structural crisis emerges not from wage depression or labor surplus independently, but from their simultaneous occurrence — consistent with the threshold mechanism described in Wage Labour and Capital.

#### Data Generating Process

```{r}
set.seed(123)
```

```{r}
N_pilot <- 10
T_pilot <- 200
```

```{r}
# labor_surplus parameters
K_l <- 1
r_l <- .05
t0_l  <- 50
A <- .2
P <- 5
delta_l <- .43
```

```{r}
# wage_gap parameters
K_w <- 1
r_base <- .03
alpha <- .5
t0_w <- 60
```

```{r}
# decay parameter
delta_c <- 0.43
```

```{r}
# Initialize storage
labor_surplus <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
wage_gap <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
CP_l <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
CP_w <- matrix(NA, nrow = T_pilot , ncol = N_pilot)
```


#### Generate wage gap and labor surplus
Labor surplus has to be generated before wage gap because the *growth rate* of the logistic trend of the wage gap depends on labor surplus

1.    Labor surplus : it represents the excess of workers relative to available job

$$
labor\_surplus (t) = \frac{K_l}{1+e^{-r_l(t-t_{0,l})}} + A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P} + \epsilon
$$

```{r}
for (n in 1:N_pilot) {
  for (t in 1:T_pilot) {
    
    logistic_trend <- K_l/(1+exp(-r_l*(t-t0_l)))
    oscillation <- A * exp(-delta_l*t) * sin((2*t*pi)/P)
    epsilon <- rnorm(1, mean = 0, sd = .02)
    
    labor_surplus[t, n] <- logistic_trend + oscillation + epsilon
  }
}
```

```{r}
head(labor_surplus)
```


2.    Wage gap : it measures the distance between the wage and the subsistence cost.


$$
wage\_gap(t) = \frac{K_w}{1+e^{-r_w(t-t_{0,w})}}
$$
*growth rate*: 

$$
r_w = r_{base,w} + \alpha(labor\_surplus_{t-1})
$$

```{r}
for (n in 1:N_pilot) {
  for (t in 1:T_pilot) {
    
    if (t == 1) {r_w <- r_base} else {r_w <- r_base + alpha*(labor_surplus[t-1, n])}
    
    wage_gap[t,n]<- K_w/(1+exp(-r_w*(t-t0_w)))
  }
}
```

```{r}
head(wage_gap)
```

The logistic formula of the wage gap is deterministic given the growth rate $r_w$ and the time $t$. 
At $t=1$, $r_w = r_{base}$ because there is no labor surplus before $t=1$. Therefore, the wage gap at $t=1$ is identical for all societies, meaning that all societies start with the same structural conditions but diverge as the market history evolves. 

#### Generate cumulative wage gap and cumulative labor surplus
Weights are assigned inversely to chronological distance, prioritizing recent events over earlier ones via an exponential decay factor.

$$
CP\_L(t) = \sum^{t-1}_{k} L_{t-k}\cdot e^{-\delta_c\cdot k}
$$

$$
CP\_W(t) = \sum^{t-1}_{k} W_{t-k}\cdot e^{-\delta_c\cdot k}
$$

```{r}
for (n in 1:N_pilot){
  for (t in 1:T_pilot){
    
    # cumulative pressure is only meaningful at t=2
    if (t==1){
      CP_l[t, n] <- 0
      CP_w[t, n] <- 0
    }
    else {
      
      cpsum_l <- 0
      cpsum_w <- 0
      
      for (k in 1:(t-1)){
        cpsum_l <- cpsum_l + labor_surplus[t-k, n] * exp(-delta_c*k)
        cpsum_w <- cpsum_w + wage_gap[t-k, n] * exp(-delta_c*k)
      }
      
      CP_l[t, n] <- cpsum_l
      CP_w[t, n] <- cpsum_w
    }
  }
  }
   
```

```{r}
head(CP_l)
```

```{r}
head(CP_w)
```

-   The identical wage gap at $t=2$ across all societies can be explain by the identical wage gap at $t=1$. At $t=2$, CP_w looks back one period, meaning it only uses data in $t=1$.

```{r}
plot(1:T_pilot, CP_l[,1], type = "l", 
     xlab = "decades", ylab = "Cumulative pressure",
     main = "cumulative labor surplus and wage gap for society 1")
lines(1:T_pilot, CP_w[,1],col ="red")
legend("topleft", legend = c("CP_l", "CP_w"), col = c("black", "red"), lty = 1)
```

**Phases of the labor surplus:**

"Wages will now rise, now fall,according to the relation of supply and demand, according as competition shapes itself between the buyers of labour-power, the capitalists, and the sellers of labour-power, the workers. "

-   *Phase 1: * Labor surplus is small and pre-inflection, driven by the logistic trend near zero plus random noise $\epsilon$. This represents the early **reserve army of labor** described by Marx — present but not yet structurally significant. 

-   *Phase 2: * New sectors emerge, temporarily increasing labor demand and absorbing surplus workers, which keeps the wage gap low. This is represented by the dampened oscillation: $A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P}$ , when $sin\frac{2t\pi}{P}$ is negative: the oscillation pulls labor surplus below the trend.

-   *Phase 3: * Labor surplus increases during the accumulation phase. During this phase, new technologies emerges and the capital grows. This leads to overproduction: more products in the market, the price falls, the profits collapse and profits collapse and workers are displaced by new technologies, swelling the reserve army. This is captured by $A e^{{-\delta_l}t}\cdot sin\frac{2t\pi}{P}$ , when $sin\frac{2t\pi}{P}$ is positive: the oscillation pushes above the trend. With each peak higher than the last as the logistic trend rises underneath.

```{r}
n <- 40
trend_l <- K_l / (1 + exp(-r_l * (10:n - t0_l)))

# Main plot
plot(10:n, labor_surplus[10:n, 1], type = "n",
     xlab = "Decades", ylab = "Labor Surplus",
     main = "Labour Surplus with Logistic Trend - Society 1")


# Add lines on top of shading
lines(10:n, labor_surplus[10:n, 1], col = "#00070d", lwd = 1.5)
lines(10:n, trend_l, col = "#E24B4A", lty = 2, lwd = 2)

# Legend
legend("topleft",
       legend = c("Labor surplus", "Logistic trend"),
       col = c("#00070d", "#E24B4A"),
       lty = c(1, 2, 1, 1), lwd = 2)
```

The plot below represent the Phase 2 and Phase 3 of the labor surplus from decade 10 to decade 40. The dotted line is the **logistic trend**: Phase 2 is the labor surplus below the trend, and Phase 3 is the labor surplus above the trend. 

Phase 2 and 3 are repeated over T, each cycle leaves more residual labor surplus than the previous, then the cumulative reserve army grows (logistic trend rises), the growing labor surplus feeds to the growth rate of the wage gap equation $r_w = r_{base,w} + \alpha(labor\_surplus_{t-1})$, and the wage gap logistic curve steepens until it reaches its inflection point $t_{0,w}$

From Phase 3 to Phase 2, there is a recovery, when new sectors appear (increase of the labor supply). But each cycle (Phase 2 + Phase 3 + Recovery ) shows in amplitude and period. These differences are explained by the **anarchic movement of the capital**: 
